### 2016-04-11 doc/publications/plos16/plos16.tex
paper sections

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is supervised learning, i.e. given a phenotype of clinical label determine the most associated features. 
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
Processing the `wide' genomic data poses difficulties for machine learning applications as having more features than samples causes the model to not generalise well and overfit the data. 
This is known as the `curse of dimensionality'~\cite{Bauer2014}. 
Random Forests (RF) as an ensemble of weak learners overcomes this issue and is hence ideally suited to this application. 
Furthermore, RF are able to capture interactions between features, which is of importance for modelling complex polygenic diseases.

@@

We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and `wide' data. Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the fist section we

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}
\subsection*{VariantSpark.}

@@

\subsection*{Synthetic data demonstrating that RF can deal with extreme dimensionality}
Detailing random association between features and labels
and whatever rob did

\subsection*{CursedForest copes with the curse of dimensionality}
\label{comp}
Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and time/mem)

@@

%\subsection*{CursedForest copes with the curse of dimensionality}
%With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
%These

>>>>

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is supervised learning, i.e. given a phenotype of clinical label determine the most associated features. 
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
Processing the `wide' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to not generalise well and overfit the data. 
This is known as the `curse of dimensionality'~\cite{Bauer2014}. 
Random Forests (RF) as an ensemble of weak learners overcomes this issue and is hence ideally suited to this application. 
Furthermore, RF are able to capture interactions between features, which is of importance for modelling complex polygenic diseases.

@@

We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and `wide' data. Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the first section we

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}

\subsection*{VariantSpark.}

@@

\section*{Random Forests converges on wide data}

As discussed in the Introduction, random forests are robust against overfitting and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on the cardinality of
the set $S$ of  ``strong predictors'' rather than on the number of variables $p$. That is, given a function $y=f(S)$
depending on a set of variables $S$, a random forest will converge to the true function $y$ with a rate that depends on
the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small compared to the total
number of variables, then the rate of convergence will be quicker.

To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a much larger feature vector $p$ per sample.  
The function, shown in Figure  \ref{figure:synth}a, is 
defined on the first 5 predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b)
and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
%TODO also how is the added noise playing into this ?
The random forest model recovers variables 1 and 4 as the two
most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

% Important ?
%In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests can recover
%the variables defining a functional relationship in the presence of a substantial number of noise variables. See the
%supplementary information for full details. It is apparently that some of the model parameters, such
%as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have to be altered
%from their default value when there are a large number of predictors.

%It is likely that in the genomic setting, variables will have a monotonic influence on the outcome. 
%We have taken \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.
%The function, shown in figure  \ref{figure:tree_function.png}, is 
%defined on the first 5 predictor variables. We have added Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png})
%and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
%The random forest model recovers variables 1 and 4 as the two
%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

%It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
%the significant varaiables if the functional relationship is simple.

%TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers; similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?  
\begin{figure}[tbhp]
\begin{tabular}{ll}
a)& b)\\
 \includegraphics[totalheight=6cm]{./figs/tree_function.png}&
  \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
 \end{tabular}
 \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$ variable with added Gaussian
   noise (in red).}
  \label{figure:synth}
\end{figure}

\subsection*{CursedForest processes genome wide variant data}
\label{comp}
Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and time/mem)

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different Machine Learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime (s)} & \bf{Memory} \\
\hline
X\% & CursedForest & &  &\\ \hline
          & RF(Spark ML) & &  &\\ \hline
          &  RF(R) & &  &\\ \hline
          &  LR & &  &\\ \hline
XX\% & CursedForest & &  &\\ \hline
          & RF(Spark ML) & &  &\\ \hline
          &  RF(R) & &  &\\ \hline
          &  LR & &  &\\ \hline

100\% & CursedForest & & & \\ \hline
          & RF(Spark ML) & &  &\\ \hline
          &  RF(R) & &  &\\ \hline
          &  LR & &  &\\ \hline
          & kmeans(Spark ML) & & & \\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
%\end{adjustwidth}
\end{table}

\subsection*{Random forest is sensitive to tree depht}

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Sensitivity analysis on parameter choice.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Mtry}  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
\hline
&&&&&\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table2}
%\end{adjustwidth}
\end{table}
